---
title: "431 Class 22"
author: "thomaselove.github.io/431"
date: "2021-11-11"
output:
  beamer_presentation:
    theme: "Madrid"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_height: 5.5
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 55)
```

## Today's Agenda: Some Loose Ends

- Working with the Favorite Movies data
- When is a complete-case analysis reasonable? MCAR!
- ANOVA
    - Assessing Assumptions with Data Visualizations
    - Dealing with Multiple Comparisons via the Holm approach 
    - Kruskal-Wallis rank-based alternative
- Wilcoxon rank-based procedures for comparing pseudo-medians
    - in paired samples (Wilcoxon signed rank)
    - in independent samples (Wilcoxon rank sum)

## Today's R Setup, Part 1

```{r, message = FALSE}
library(knitr); library(magrittr); library(naniar)
library(broom); library(patchwork)

library(googlesheets4)
source("data/Love-boost.R")
```

```{r}
library(janitor)
```

## Today's R Setup, Part 2

```{r}
library(tidyverse)
```

- The 8 core tidyverse packages are listed every time you load the tidyverse (unless you tell R not to show messages.)

```{r}
theme_set(theme_bw())
```


## Favorite Movies

```{r, message = FALSE, cache = TRUE}
url_raw <- "https://docs.google.com/spreadsheets/d/1t4668vGN-97284uWbH154LQHmj8p90hkIlafPOjnIcQ/edit"

movie_raw <- read_sheet(url_raw, na = c("", "NA")) %>%
    arrange(film_id) %>%
    clean_names()

dim(movie_raw)

write_rds(movie_raw, "data/movie_raw.Rds")
```

## Or, more simply, since we have the `.Rds` file

```{r}
movie_raw <- read_rds("data/movie_raw.Rds")

dim(movie_raw)
```

## Missing Data?

```{r}
miss_var_summary(movie_raw)
```

## Picking Out A Few Variables

```{r}
movie1 <- movie_raw %>%
    select(film_id, film, imdb_stars, mpa)

head(movie1) %>% kable()
```

## How many films in each `mpa` category?

```{r}
movie1 %>% tabyl(mpa)
```

Let's look at the three categories with at least 30 films.

```{r}
movie1 <- movie1 %>%
    filter(mpa %in% c("PG", "PG-13", "R"))

nrow(movie1)
```

## ANOVA assumptions

1. Approximately Normal distribution of outcome within each group
    - Can use a rank-based alternative (Kruskal-Wallis) if there is a serious concern, or consider a transformation of the outcome
2. Equal population variance of outcome within each group (extension of pooled t)
3. Independently drawn samples of the outcome within each group

## Plot IMDB Stars by MPA Rating

```{r, fig.height = 3.5}
ggplot(movie1, aes(x = mpa, y = imdb_stars)) +
    geom_violin(aes(fill = mpa)) +
    geom_boxplot(width = 0.3, outlier.size = 3) +
    stat_summary(aes(fill = mpa), fun = mean, 
                 geom="point", pch = 21, size = 4) + 
    guides(fill = "none") + coord_flip()
```

## Numerical Summary of IMDB Stars by MPA Rating

```{r, message = FALSE}
mosaic::favstats(imdb_stars ~ mpa, data = movie1) %>%
    kable(digits = 2)
```

## Analysis of Variance of IMDB Stars by MPA

```{r}
mod1 <- lm(imdb_stars ~ mpa, data = movie1)
anova(mod1) %>% kable(digits = 3)
tidy(mod1, conf.int = TRUE, conf.level = 0.90) %>%
    select(term, estimate, std.error, conf.low, conf.high) %>%
    kable(digits = 3)
```

## Bonferroni pairwise comparisons for mpa groups

```{r}
movie1 %$%
    pairwise.t.test(imdb_stars, mpa, 
                    p.adjust.method = "bonferroni")
```

## Holm pairwise comparisons across mpa groups

- Works well even with an unbalanced design so long as ANOVA assumptions hold.
- Not as conservative as Bonferroni, but uniformly more powerful.

```{r}
movie1 %$%
    pairwise.t.test(imdb_stars, mpa, 
                    p.adjust.method = "holm")
```

## Tukey HSD set of Pairwise Comparisons

```{r}
tukey1 <- movie1 %$% 
    TukeyHSD(aov(imdb_stars ~ mpa), 
             ordered = TRUE, conf.level = 0.90)

tidy(tukey1) %>% kable(digits = 3)
```

## Tukey HSD set of Pairwise Comparisons

```{r, height = 5}
plot(tukey1)
```

## Residual Analysis for our ANOVA model

```{r}
par(mfrow = c(2,2)); plot(mod1); par(mfrow = c(1,1))
```

## Kruskal-Wallis Rank-Based ANOVA Approach

```{r}
kruskal.test(imdb_stars ~ mpa, data = movie1)
```

- No longer comparing means, and no confidence intervals here.
- No straightforward decision about what to do about pairwise comparisons, other than Holm-based comparisons based on Wilcoxon rank sum tests.
- Speaking of Wilcoxon rank-based tests...

## A New Question...

Are students in 431 more likely to have seen movies that were nominated for Academy Awards?

- How many of the 115 movies received Oscar nominations?

```{r}
movie2 <- movie_raw %>%
    select(film_id, film, x431_pct, oscar_noms) %>%
    mutate(oscar_n = factor(
        ifelse(oscar_noms > 0, "Yes", "No")))
```

```{r}
movie2 %>% tabyl(oscar_n)
```

## % of students who've seen film, by Oscar Status?

```{r, fig.height = 3.5}
ggplot(movie2, aes(x = oscar_n, y = x431_pct)) +
    geom_violin(aes(fill = oscar_n)) +
    geom_boxplot(width = 0.2, outlier.size = 3) +
    stat_summary(aes(fill = oscar_n), fun = mean, 
                 geom="point", pch = 21, size = 4) + 
    guides(fill = "none") + coord_flip() +
    scale_fill_viridis_d(alpha = 0.5)
```

## Comparing Two Population Means

Does this design involve paired samples, or independent samples? Why?

```{r}
mosaic::favstats(x431_pct ~ oscar_n, data = movie2) %>%
    kable(digits = 2)
```

## Comparing 431 % by Oscar Nomination Status

We have four different approaches. Two are based on the t distribution.

```{r}
# pooled t 
compA <- t.test(x431_pct ~ oscar_n, data = movie2,
                var.equal = TRUE, conf.level = 0.90)

tidy(compA) %>% 
    select(estimate1, estimate2, estimate, 
           conf.low, conf.high) %>% kable(dig = 2)
```

## Comparing 431 % by Oscar Nomination Status

Here's the second one based on the t distribution.

```{r}
# Welch's t
compB <- t.test(x431_pct ~ oscar_n, data = movie2,
                conf.level = 0.90)

tidy(compB) %>% 
    select(estimate1, estimate2, estimate, 
           conf.low, conf.high) %>% kable(dig = 2)
```

## Comparing 431 % by Oscar Nomination Status

This strategy doesn't use the t distribution.

```{r}
# bootstrap
set.seed(431)
compC <- movie2 %$% bootdif(x431_pct, oscar_n, 
                            conf.level = 0.90)

compC
```

## Comparing 431 % by Oscar Nomination Status

What if we rank the observations from low to high in each group, then compare the results?

```{r}
# Wilcoxon-Mann-Whitney rank sum with continuity correction
compD <- wilcox.test(x431_pct ~ oscar_n, data = movie2,
                     conf.int = TRUE, conf.level = 0.90)

tidy(compD) %>% select(estimate, conf.low, conf.high) %>%
    kable(digits = 2)
```

## Interpreting the Rank Sum Test

The Wilcoxon-Mann-Whitney rank sum test (also called the Mann-Whitney U test and lots of other things) tests the null hypothesis that if we randomly select X and Y from the two populations of interest, the probability of X being greater than Y is the same as the probability of Y being greater than X.

- The "Pseudomedian" is sometimes referred to as the Hodges-Lehmann estimate. It is the median of all possible differences between an observation in the first sample and an observation in the second sample.

To write up a Wilcoxon rank sum result, we usually suggest specifying the two medians directly, and then describing the *p* value or (less commonly) a confidence interval.

The Wilcoxon-Mann-Whitney test requires only that the data be ordinal, and this reduces the influence of outliers. It doesn't test the same thing as the t test (or bootstrap) however.

## Comparing 431 % by Oscar Nomination Status

**Four Comparisons for these Independent Samples**

Method | Yes - No Est. | 90% CI | Statistic
:------: | :---------: | :-------: | :--------:
Pooled t | 16.73 | (9.58, 23.88) | Mean
Welch's t | 16.73 | (9.79, 23.66) | Mean
Bootstrap | 16.73 | (9.97, 23.42) | Mean
Rank Sum | 17 | (10, 24) | "Pseudo-median"

### Numerical Summaries from the Data

```{r, echo = FALSE}
movie2 %>% group_by(oscar_n) %>% 
    summarize(n = n(), mean = mean(x431_pct), 
              median = median(x431_pct), sd = sd(x431_pct)) %>%
    kable(digits = 3)
```

## Comparing A New Outcome under 2 Conditions

Suppose we want to compare the percentage of **critics** who recommend a film to the percentage of **the general audience** who recommend the film. Each film has:

- `rt_critics` = from Rotten Tomatoes: percentage of critics who recommend the film (sample mean across 113 films was `r round_half_up(mean(movie_raw$rt_critics, na.rm = TRUE),1)`)
- `rt_audience` = from Rotten Tomatoes: percentage of audience who recommend the film (sample mean across 114 films was `r round_half_up(mean(movie_raw$rt_audience, na.rm = TRUE),1)`)

Can we compare the difference between the means (at least for the 113 films with data on each variable) appropriately?

What's the design we have here?

## Creating our Third Data Set

```{r}
movie3 <- movie_raw %>% 
    filter(complete.cases(rt_critics, rt_audience)) %>%
    select(film_id, film, rt_critics, rt_audience) 

dim(movie3)
```

We lost two films (Farewell My Concubine and Jab We Met) which didn't have information on each of our variables.

## The `movie3` data

```{r}
movie3 <- movie3 %>%
    mutate(diff = rt_critics - rt_audience)

head(movie3)
```

We can develop paired differences in this paired samples setting. What do those differences look like?

---

```{r, echo = FALSE}
p1 <- ggplot(movie3, aes(sample = diff)) +
  geom_qq(col = "magenta2") + geom_qq_line(col = "navy") + 
  theme(aspect.ratio = 1) + 
  labs(title = "Normal Q-Q: Critics - Audience")

p2 <- ggplot(movie3, aes(x = diff)) +
  geom_histogram(aes(y = stat(density)), 
                 bins = 15, fill = "magenta2", col = "white") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(movie3$diff), 
                            sd = sd(movie3$diff)),
                col = "navy", lwd = 1.5) +
  labs(title = "Density + Normal: Critics - Audience")

p3 <- ggplot(movie3, aes(x = diff, y = "")) +
  geom_boxplot(fill = "magenta2", outlier.color = "magenta2") + 
  labs(title = "Boxplot: Critics - Audience", y = "")

p1 + (p2 / p3 + plot_layout(heights = c(4,1)))

mosaic::favstats(~ diff, data = movie3) %>% kable(digits = 1)
```

## Paired T approach

```{r}
compX <- movie3 %$% t.test(diff, conf.level = 0.90)

tidy(compX) %>% 
    select(estimate, conf.low, conf.high, p.value) %>% 
    kable(dig = 2)
```

## Bootstrap approach

```{r}
set.seed(431431)
compY <- movie3 %$% 
    Hmisc::smean.cl.boot(diff, conf.int = 0.90)

compY
```

## Wilcoxon Signed Rank approach

```{r}
compZ <- movie3 %$% 
    wilcox.test(diff, conf.int = 0.90)

tidy(compZ) %>%
    select(estimate, conf.low, conf.high, p.value) %>% 
    kable(dig = 2)
```
## Comparing % Recommend by Critics vs. Audiences

**Three Comparisons for these Paired Samples**

Method | Crit - Aud Est. | 90% CI | Statistic
:------: | :---------: | :-------: | :--------:
t | -1.95 | (-4.17, 0.28) | Mean difference
Bootstrap | -1.95 | (-4.25, 0.17) | Mean difference
Signed Rank | -0.5 | (-3, 1.5) | "Pseudo-median" difference

### Summarizing the Data

```{r, echo = FALSE}
summ1 <- mosaic::favstats(~ rt_critics, data = movie3) %>% 
    mutate(group = "critics")
summ2 <- mosaic::favstats(~ rt_audience, data = movie3) %>% 
    mutate(group = "audience")
summ3 <- mosaic::favstats(~ diff, data = movie3) %>% 
    mutate(group = "c-a diffs")
bind_rows(summ1, summ2, summ3) %>%
    select(group, mean, sd, median, n, min, max) %>% kable(dig = 2)
```
