---
title: "431 Class 08"
author: "thomaselove.github.io/431"
date: "2021-09-16"
output:
  beamer_presentation:
    theme: "Madrid"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_height: 5.5
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 55)
```

## Today's Agenda

- Building Visualizations to Compare Two Distributions
  - Confidence Intervals for a Difference between Means
- Building Visualizations to Compare > 2 Distributions

## Today's R Packages

```{r, message = FALSE}
library(broom) # for tidying up output
library(janitor)
library(knitr)
library(magrittr)
library(naniar)
library(patchwork)
library(readxl) # new today, read in .xls or .xlsx files
library(tidyverse)

theme_set(theme_bw())
```

## Today's Data

Today, we'll use an Excel file (.xls, rather than .csv) to import the `dm1000` data.

```{r}
dm1000 <- read_excel("data/dm_1000.xls") %>%
  clean_names() %>%
  mutate(across(where(is.character), as_factor)) %>%
  mutate(subject = as.character(subject))
```

- There are also functions called `read_xls()` and `read_xlsx()` available in the `readxl` package.

## The `dm1000` tibble

```{r, echo = FALSE}
dm1000
```

# Comparing Two Distributions

## LDL cholesterol and statin prescription?

```{r, fig.height = 4}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = ldl, fill = factor(statin))) +
  geom_histogram(bins = 15, col = "white") +
  facet_grid(statin ~ ., labeller = "label_both") +
  guides(fill = "none")
```

## Faceting

It's very useful to split data into groups and plot each group separately to make comparisons across the groups. We can then draw those subplots side by side.

We have two main tools: `facet_wrap()` and `facet_grid()`

- `facet_wrap(~ grp1)` to obtain plots within each `grp1` arranged into horizontal subpanels and wrapping around, like words on a page.
- `facet_grid(grp1 ~ .)` to obtain plots within each `grp1` arranged vertically (vertical subpanels)
- `facet_grid(grp1 ~ grp2)` to obtain plots within each combination of `grp1` and `grp2` with vertical and horizontal subpanels.

## Using `facet_wrap()`

```{r, fig.height = 3.5}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = ldl, fill = factor(statin))) +
  geom_histogram(binwidth = 10, col = "white") +
  facet_wrap(~ statin, labeller = "label_both") +
  guides(fill = "none")
```

## Using `facet_grid()`

```{r, fig.height = 3.5}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = ldl, fill = factor(statin))) +
  geom_histogram(binwidth = 10, col = "white") +
  facet_grid(statin ~ ., labeller = "label_both") +
  guides(fill = "none")
```

## Using `facet_grid()` with two groupings

```{r, fig.height = 3.5}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = ldl, fill = factor(statin))) +
  geom_histogram(binwidth = 10, col = "white") +
  facet_grid(statin ~ residence, labeller = "label_both") +
  guides(fill = "none")
```

## Comparison of densities (ignores relative frequency)

```{r, fig.height = 4}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = ldl, fill = factor(statin))) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +
  labs(fill = "Statin")
```

## Numerical Summaries comparing Two Groups

```{r}
dm1000 %>% filter(complete.cases(statin, ldl)) %>% 
  group_by(statin) %>%
  summarize(n = n(), min = min(ldl), med = median(ldl), 
            max = max(ldl), mean = mean(ldl), 
            sd = sd(ldl)) %>%
  kable(digits = 2)
```

- What is the difference in mean(LDL) between the two samples?

## Using `favstats` to compare LDL by Statin group

```{r, message = FALSE}
dm1000 %$% 
  mosaic::favstats(ldl ~ statin)
```

We would have obtained the same result with:

```{r, eval = FALSE}
mosaic::favstats(ldl ~ statin, data = dm1000)
```

## Comparison Boxplot with Violins (LDL and statin)

Here's a first attempt...

```{r, fig.height = 3}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = factor(statin), y = ldl)) +
  geom_violin() +
  geom_boxplot(aes(fill = factor(statin)), width = 0.3) + 
  labs(x = "Statin prescription (0 = no, 1 = yes)", 
       y = "LDL in mg/dl", fill = "Statin")
```

## Try 2: Boxplot with Violins for LDL and statin

```{r, fig.height = 3}
dm1000 %>% filter(complete.cases(ldl, statin)) %>%
  ggplot(data = ., aes(x = factor(statin), y = ldl)) +
  geom_violin(aes(fill = factor(statin))) +
  geom_boxplot(width = 0.3, outlier.size = 3) + 
  coord_flip() +
  guides(fill = "none") +
  labs(x = "Statin prescription (0 = no, 1 = yes)", 
       y = "LDL in mg/dl")
```

## Setting Up Third Try

```{r}
dm_for_boxplot <- dm1000 %>%
  filter(complete.cases(statin, ldl)) %>%
  mutate(statin_f = fct_recode(factor(statin),
                               "No Statin" = "0",
                               "Statin" = "1")) %>%
  select(subject, ldl, statin_f, statin)

head(dm_for_boxplot, 3) # print first three rows
```

## Third Try on Boxplot for LDL by Statin Use

```{r, fig.height = 3}
ggplot(data = dm_for_boxplot, aes(x = statin_f, y = ldl)) +
  geom_violin(aes(fill = statin_f)) +
  geom_boxplot(width = 0.3, outlier.size = 3) + 
  coord_flip() + guides(fill = "none") +
  scale_fill_viridis_d(alpha = 0.6) +
  labs(x = "Statin prescription", 
       y = "LDL in mg/dl")
```

## 95% confidence interval for difference between population mean LDL WITH statin and population mean LDL WITHOUT statin

**If** we are willing to assume that LDL follows a Normal distribution in each statin group, then we can use the following linear model with one predictor.

```{r}
m2 <- lm(ldl ~ statin, data = dm1000)
tidy(m2, conf.int = TRUE, conf.level = 0.95) %>% 
  select(term, estimate, conf.low, conf.high) %>%
  kable(digits = 3)
```

## Alternative Approach to get same result

```{r}
tt <- t.test(ldl ~ statin, data = dm1000, 
       var.equal = TRUE, conf.level = 0.95)
tidy(tt) %>% select(estimate, conf.low, conf.high) %>%
  kable(digits = 3)
```


## 95% confidence interval for difference between population mean LDL WITH statin and population mean LDL WITHOUT statin

If we are not willing to assume a Normal distribution for LDL in either the statin or the "no statin" group, then we could use a bootstrap approach.

```{r}
source("data/Love-boost.R")
set.seed(20210914)
dm1000 %$% bootdif(y = ldl, 
                   g = factor(statin), 
                   conf.level = 0.95)
```

## The `bootdif` function (from `Love-boost.R`)

```{r}
`bootdif` <-
  function(y, g, conf.level=0.95, B.reps = 2000) {
    lowq = (1 - conf.level)/2
    g <- as.factor(g)
    a <- attr(Hmisc::smean.cl.boot(y[g==levels(g)[1]], 
                          B=B.reps, reps=TRUE),'reps')
    b <- attr(Hmisc::smean.cl.boot(y[g==levels(g)[2]], 
                          B=B.reps, reps=TRUE),'reps')
    meandif <- diff(tapply(y, g, mean, na.rm=TRUE))
    a.b <- quantile(b-a, c(lowq,1-lowq))
    res <- c(meandif, a.b)
    names(res) <- c('Mean Difference',lowq, 1-lowq)
    res
  }
```

## 95% confidence intervals for $\mu_{No Statin} - \mu_{Statin}$

Approach | Estimate | 95% CI | Normality Assumption?
--------: | -------: | -----: | --------
linear model | 7.00 | (0.83, 13.17) | Yes
bootstrap | 7.00 | (0.90, 12.96) | No

Assumptions these intervals share:

- random samples from the populations of interest
- independent samples (samples aren't paired or matched)

Additional assumptions for linear model:

- Normal distribution in each group (statin and "no statin")
- variance in each group (statin and "no statin") is equal


# Comparing Multiple (more than 2) Batches of Data

## Stratify the subjects by primary `insurance`?

```{r}
dm1000 %>% count(insurance) %>%
  mutate(pct = 100*n/sum(n)) %>%
  kable(digits = 1)
```


## Let's look at `n_income`

```{r, echo = FALSE, warning = FALSE, message = FALSE}
p1 <- ggplot(dm1000, aes(sample = n_income)) +
  geom_qq(col = "slateblue") + 
  geom_qq_line(col = "magenta") + 
  theme(aspect.ratio = 1) + 
  labs(title = "Normal Q-Q plot: dm1000 n_income")

p2 <- ggplot(dm1000, aes(x = n_income)) +
  geom_histogram(aes(y = stat(density)), 
                 bins = 20, fill = "slateblue", col = "cyan") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(dm1000$n_income, na.rm = TRUE), 
                            sd = sd(dm1000$n_income, na.rm = TRUE)),
                col = "magenta", lwd = 1.5) +
  labs(title = "Density Function: dm1000 n_income")

p3 <- ggplot(dm1000, aes(x = n_income, y = "")) +
  geom_boxplot(fill = "slateblue", outlier.color = "slateblue") + 
  labs(title = "Boxplot: dm1000 n_income", y = "")

p1 + (p2 / p3 + plot_layout(heights = c(4,1)))

mosaic::favstats(~ n_income, data = dm1000) %>% kable(digits = 0)
```

## Compare `n_income` by insurance: Boxplot?

```{r, fig.height = 4}
ggplot(dm1000, aes(x = insurance, y = n_income)) +
  geom_boxplot()
```


## Build a better boxplot?

What am I doing in each line of this code?

```{r, eval = FALSE}
dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income/1000)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  labs(y = "Neighborhood Income (in $1000s)",
       x = "Primary Insurance Category",
       title = "dm1000 Income by Insurance Type")
```

- Result on next slide...

## How does `n_income` vary by `insurance` in `dm1000`?

```{r, echo = FALSE}
dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income/1000)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  labs(y = "Neighborhood Income (in $1000s)",
       x = "Primary Insurance Category",
       title = "dm1000 Income by Insurance Type")
```

## Faceted Histograms of `n_income` by `insurance`

```{r, fig.height = 4}
ggplot(data = dm1000, aes(x = n_income)) +
  geom_histogram(binwidth = 2500) +
  facet_wrap(~ insurance)
```

## Improving the Histograms (result)

```{r, echo = FALSE}
dm1000 %>% filter(complete.cases(n_income, insurance)) %>%
ggplot(data = ., aes(x = n_income, fill = insurance)) +
  geom_histogram(binwidth = 2500, col = "navy") +
  scale_x_continuous(
    breaks = c(5000, 25000, 50000, 75000, 100000)) +
  guides(fill = "none") +
  facet_wrap(~ insurance) +
  labs(title = "Neighborhood Income, in $")
```

## Improving the Histograms (code)

```{r, eval = FALSE}
dm1000 %>% filter(complete.cases(n_income, insurance)) %>%
ggplot(data = ., aes(x = n_income, fill = insurance)) +
  geom_histogram(binwidth = 2500, col = "navy") +
  scale_x_continuous(
    breaks = c(5000, 25000, 50000, 75000, 100000)) +
  guides(fill = "none") +
  facet_wrap(~ insurance) +
  labs(title = "Neighborhood Income, in $")
```

## Comparing Densities of `n_income` by `insurance`

```{r, fig.height = 3.5}
dm1000 %>% filter(complete.cases(n_income, insurance)) %>%
  ggplot(data = ., aes(x = n_income, fill = insurance)) +
  geom_density() +
  scale_fill_viridis_d(alpha = 0.3)
```

## Using a Ridgeline Plot to Compare Densities

```{r, fig.height = 3.5}
dm1000 %>% filter(complete.cases(n_income, insurance)) %>%
  ggplot(data = ., aes(x = n_income, y = insurance, 
                       fill = insurance)) +
  geom_density_ridges(alpha = 0.5) +
  labs(title = "Neighborhood Income, in $")
```

## Sample Summaries of `n_income`

```{r}
dm1000 %$% mosaic::favstats(n_income ~ insurance)
```


## Comparing the Means of `n_income`

```{r}
m1 <- lm(n_income ~ insurance, data = dm1000)

tidy(m1) %>% select(term, estimate) %>% kable(digits = 2)
```

- What is m1's estimated `n_income` for each of the insurance groups?

## `m1`: Estimated `n_income` for each insurance

```
n_income = 33150.74 + 7564.28 Commercial  
                   + 732.76 Medicare + 4724.00 Uninsured
```

Insurance | Estimated `n_income` 
--------: | --------: 
Commercial | 33150.74 + 7564.28 = 40715.02 
Medicare | 33150.74 + 732.76 = 33883.50
Uninsured | 33150.74 + 4724.00 = 37874.74
Medicaid | 33150.74

## `m1`: Estimated `n_income` for each insurance

```
n_income = 33150.74 + 7564.28 Commercial  
                   + 732.76 Medicare + 4724.00 Uninsured
```

Insurance | `m1` est. | Sample mean(`n_income`)
--------: | :--------: | :--------:
Commercial | 40715.0 | 40715.0
Medicare | 33883.5 | 33883.5
Uninsured | 37874.7 | 33874.7
Medicaid | 33150.7 | 33150.7

## Model `m1` coefficients

```{r}
tidy(m1, conf.int = TRUE, conf.level = 0.95) %>%
  select(term, estimate, conf.low, conf.high) %>%
  kable(digits = 2)
```

- 95% CI for pop. mean `n_income` among adults with Medicaid is (31097, 35205)

## Model `m1` coefficients

```{r}
tidy(m1, conf.int = TRUE, conf.level = 0.95) %>%
  select(term, estimate, conf.low, conf.high) %>%
  kable(digits = 2)
```

- 95% CI for Commercial - Medicaid is (4237, 10891)
- What about Medicare - Medicaid or Uninsured - Medicaid?

## Comparing `n_income` across `insurance` groups

```{r, message = FALSE}
mosaic::favstats(n_income ~ insurance, data = dm1000)
```

- Does a comparison of means make sense here?
- Would it give us the same conclusions as comparing medians?

## Replot on logarithmic (base 10) scale?

```{r, echo = FALSE}
dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_y_continuous(trans = "log10") +
  labs(y = "Neighborhood Income (in $, log10 scale)",
       x = "Primary Insurance Category",
       title = "dm1000 Income by Insurance Type")
```

## Code for plot on previous slide

```{r, eval = FALSE}
dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_y_continuous(trans = "log10") +
  labs(y = "Neighborhood Income (in $, log10 scale)",
       x = "Primary Insurance Category",
       title = "dm1000 Income by Insurance Type")
```

- Could also use `scale_y_log10()`

## Does which logarithmic scale you pick matter?

```{r, echo = FALSE}
p1 <- dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_y_continuous(trans = "log") +
  labs(y = "Neighborhood Income (in $, log, base e scale)",
       x = "Primary Insurance Category",
       title = "Natural Log Scale")

p2 <- dm1000 %>% filter(complete.cases(insurance, n_income)) %>%
  ggplot(data = ., aes(x = insurance, y = n_income)) +
  geom_violin(aes(fill = insurance)) +
  geom_boxplot(width = 0.3, outlier.size = 2) +
  guides(fill = "none") +
  coord_flip() +
  scale_fill_viridis_d(alpha = 0.5) +
  scale_y_continuous(trans = "log10") +
  labs(y = "Neighborhood Income (in $, log10 scale)",
       x = "Primary Insurance Category",
       title = "Log (base 10) Scale")

p1 + p2
```

## Ridgeline Plot of Densities on Log Scale

```{r, fig.height = 3.5, message = FALSE}
dm1000 %>% filter(complete.cases(n_income, insurance)) %>%
  ggplot(data = ., aes(x = n_income, y = insurance, 
                       fill = insurance)) +
  geom_density_ridges(alpha = 0.5) +
  scale_x_continuous(trans = "log10") +
  labs(title = "Neighborhood Income, log10 scale")
```

## Next Up

- Favorite Movies activity
- Correlation, Association and Scatterplots